MAP566 - Homework assignment #1

### 2024-03-01 homework_clean.Rmd
feat: method justifications for ex4

@@

>>>>

    ```{r}
    ggplot(sp500_data, aes(x = Date, y = DailyReturn)) +
      geom_histogram(stat = "identity", fill = "grey", color = "grey", bins = 15) +
      labs(title = "Daily Return of S&P 500 (2007-2010)",
           x = "Date",
           y = "Daily Return") +
      theme_minimal() +
      theme(axis.text.x = element_blank(),
            axis.title.x = element_blank())
    ```

    Under Louis Bachelier's "Theory of Speculation" assume that the returns can be sampled from a perfectly random distribution, in this case this would be a normal distribution which we can parametirise through a Gaussian model. Assume that the return is modeled by some random variable $X_r$ and that some value of that RV is denoted by $x_t$. Thus, since we first assume Bachelier's theory we can model the RV through the following Gaussian.

    $$
    X_r \sim \mathcal{N}(\mu_r, \sigma_r^2)
    $$

    Where the model parameters are the empirical mean and empirical variance expressed by:

    $$
    \mu_r = \dfrac{\sum_{i=0}^N x_i}{N} \hspace{5mm} \sigma_r^2 = \dfrac{\sum_{i=0}^N (x_i - \mu_r)}{N}
    $$

    We can then directly fit such a Gaussian model in R through the following:

    ```{r}
    fit <- fitdistr(sp500_data$DailyReturn, "normal")

    # Print the fitted parameters
    print(fit)

    # Plot the histogram of daily returns with the fitted Gaussian distribution
    hist(sp500_data$DailyReturn, breaks = 100, freq = FALSE, 
         main = "Histogram of Daily Returns with Fitted Gaussian Distribution",
         xlab = "Daily Return")
    curve(dnorm(x, mean = fit$estimate[1], sd = fit$estimate[2]), 
          col = "blue", lwd = 2, add = TRUE, yaxt = "n")
    ```

    Indeed, the it seems that Bachelier's theory does describe the price movements well. As the distribution of the returns follow quite clearly a normal distribution.

@@

    To combine the multiple Gaussians to model a single valid PDF, we will have to introduce a weight parameter which will be used between the weighted sum of Gaussian components. First, we denote a single component the following way.

    $$
    Y_r \sim \mathcal{N}(\theta_k) \hspace{5mm} \text{Where $\theta$ is a parameter vector for component $k$ defined by} \hspace{5mm} \theta_k = \left[ \mu_k, \sigma_k^2 \right]
    $$

    Furthermore, to form the mixture we introduce the weight parameter $\pi_k$ which is probabilistic and forms the weighted sum (GMM).

    $$
    Y_i = \sum_{k=1}^K \pi_k \mathcal{N}(\theta_k)
    $$

    As this is a weighted sum and we expect $Y_i$ to be modeled from a valid PDF, we have to constrain the weights such that:

    $$
    \pi_k \geq 0 \hspace{2.5mm} \land \hspace{2.5mm} \sum_{k=1}^K \pi_k = 1
    $$

    Which is the case as long as we follow that $\pi_k$ comes from a PDF defined over the categorical RVs $Z_i$ which defines the different components such that we have:

    $$
    P(Z_i = k) = \pi_k \hspace{5mm} \text{Where $Z$ is a sequence of RVs} \hspace{5mm} \{1, 2, \dots, K\}
    $$

    Thus, we observe that we first have to initialize the parameters of the model which we can do by placing the means based on a K-Means initialization and set some explicit variance which will serve as the initialization for the iterative EM-algorithm for parameter optimization over such a GMM.
